Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-01-09 Manuscript_Nater_etal.qmd
updated disc and bib

There is growing demand for biodiversity indicators from international unions, national government, local management bodies, and corporate and industry actors. Such indicators ideally represent a wide range of biodiversity's states and functions [e.g. Essential Biodiversity Variables, @pereira2013; @jetz2019], yet the development of suitable indicators for certain attributes, such as species abundance and demography, has been more difficult than for others [@schmeller2018; @waldock2022]. This is at least partially due to challenging requirements regarding spatial scales of useful biodiversity indicators. On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets [e.g. @feld2009]. On the other hand, indicators also ideally have good spatial resolution as the scales relevant for local-level management and planning are often much smaller [@stevenson2021]. This latter requirement is particularly crucial for (infrastructure) development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e. demographic rates) that is relevant for their county- or municipality-level decision making (+ refs?). Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space [e.g. @robinson2014; @horswill2019]. This is variation that needs to be accounted for in order to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].

@@

There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e. the collection of ecological data by members of the public [also called citizen science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023]. Consequently, many large-scale monitoring programmes are often limited to presence-absence or very simple count observations, making them suitable for the development of indicators of species distributions and perhaps population trends, but usually not of abundance, population dynamics, and demographic rates (+ example refs). The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol. For example, in the United States hunters participate in the collection of bands and wings from harvested American Woodcock (*Scopolax minor)* to estimate survival and age ratios [@zimmerman2010]. <!-- can we think of 1-2 internationally well-known examples? Check out MAPS (Monitoring of Avian Productivity and Survivorship) and CES (Constant Effort Sites) --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", <https://honsefugl.nina.no/Innsyn/en>). It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs. The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via GBIF. As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.

The line transect data from "Hønsefuglportalen" has been used previously for estimating abundance trends of willow ptarmigan (*Lagopus lagopus*) across Norway [e.g. @bowler2020; @nilsen2020] but large-scale estimation of demographic rates underlying these trends has thus far remained as untapped potential of the dataset. @nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses age information in line transect data alongside radio-telemetry data to estimate not just abundance, but also survival and recruitment across years. In this study, we adapt the model of @nilsen2023 to run not just on a single site but on all areas with publicly available line transect data from "Hønsefuglportalen" simultaneously. Unlike several previous studies applying integrated models for population dynamics to multiple (sub-) populations separately and comparing results [e.g. @robinson2014; @nater2023], we opt for an approach explicitly integrating across space, thus allowing for sharing of information across locations and -- in effect -- space-for-time substitution [e.g. @horswill2019; @morrison2022]. We then apply the resulting multi-area IDSM to "Hønsefuglportalen" data on willow ptarmigan to estimate population size, age-structure, survival, recruitment, and impacts of small rodent abundance across 41 reporting districts and 15 years (2007-2021) for this popular small-game species. We further embed the modelling workflow in a reproducible, semi-automated pipeline that will greatly facilitate the repeated calculation of abundance and population indicators at different spatial scales as new data is added every year.

@@

The willow ptarmigan (*Lagopus lagopus*) is a tetraoid bird with a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g. @fuglei2020]. While the species is currently listed as Least Concern (LC) both in the global and Norwegian Red List of Species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022]. While the main reason for the long-term decline in abundance remain unresolved, willow ptarmigan are considered sentinel species that are sensitive to both climate change and land use changes [@Henden2017; @Storch2007]. Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for the gyrfalcon [@Alastair2020] <!-- do we have other "important" species to add here. Arctic fox? Some other endangered beasts perhaps? -->. The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@Bowler2020]. Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @Bowler2020]. This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis and has been a central part of Fennoscandian research on grouse population dynamics for many decades [@hagen1952; @Linden1988: @Steen1985]. In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023, @Steen1985]. Across their distributional range, they are iconic species with a high cultural value, partly linked to their popularity as game species. The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies. In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem [Assessment of the Ecological Condition, @framstad2022] assessments.

@@

In this study we used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different reporting districts spanning 9 counties and 50 municipalities. Transects on which no willow ptarmigan were ever observed (i.e. species absence likely due to low habitat suitability) were dropped, resulting in 2077 transects used in the analyses.

@@

In this study we used data from years 2015 - 2020, and the total sample size across these years was 139 birds for the Aug-Jan period and 258 birds for the Feb-Jul period.

@@

@@

The highest population growth rates were estimated for areas with relatively low population densities across latitudes [@fig-paramCorrs A] but we did not find evidence for a strong association between population growth rates and population densities overall.

@@

Recruitment rates varied between `r R.min$post.sum` (area "`r R.min$area.name`") and `r R.max$post.sum` (area "`r R.max$area.name`") and displayed a spatial pattern opposite to that of annual survival, i.e. lower recruitment rates co-occurring with higher survival rates and vice-versa (@fig-pVRB; @fig-paramCorrs B). Across all areas and years, average recruitment rate was `r R.hyper.mean$post.sum`. Unlike for survival, the model predicted similar magnitudes of spatial and temporal variation (random effect SDs of `r R.sd.spatial$post.sum` and `r R.sd.temporal$post.sum`, respectively), and about twice as much residual variation (`r R.sd.residual$post.sum`).

@@

Detection decay parameters, which determine detection probability in distance sampling surveys, varied across areas from between `r Det.min$post.sum` in "`r Det.min$area.name`" to `r Det.max$post.sum` in "`r Det.max$area.name`, resulting in detection probabilities over the transect sites (truncation distance = 200 m) from `r pDet.min$post.sum` to `r pDet.max$post.sum`, respectively. The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF "Avg_detect_Map.pdf"). Variation in detection over time was modest (SF "TimeSeries_pDetect.pdf"), and among-year variation in detection decay (random effect log SD = `r Det.sd.temporal$post.sum`) was almost identical in magnitude to spatial (`r Det.sd.spatial$post.sum`) and residual (`r Det.sd.residual$post.sum`) variation.

@@

@@

@@

@@

-   Management also includes optimization of monitoring

>>>>

There is growing demand for biodiversity indicators from international unions, national government, local management bodies, and corporate and industry actors. Such indicators ideally represent a wide range of biodiversity's states and functions [e.g. Essential Biodiversity Variables, @pereira2013; @jetz2019], yet the development of suitable indicators for certain attributes, such as species abundance and demography, has been more difficult than for others [@schmeller2018; @waldock2022]. This is at least partially due to challenging requirements regarding spatial scales of useful biodiversity indicators. On one hand, indicators need to be representative at large geographic scales, for example, in the context of countries' reporting towards biodiversity targets [e.g. @feld2009]. On the other hand, indicators also ideally have good spatial resolution as the scales relevant for local-level management and planning are often much smaller [@stevenson2021]. This latter requirement is particularly crucial for (infrastructure) development strategies and for species management and conservation, both of which tend to require knowledge on species abundance and population dynamics (i.e. demographic rates) that is relevant for their county- or municipality-level decision making (+ refs?). Another reason why abundance and population indicators ideally come with good spatial resolution is that there can be substantial amounts of variation in population dynamics and life history of species across space [e.g. @robinson2014; @horswill2019]. This variation that needs to be accounted for to develop successful and sustainable strategies for area use, (harvest) management, and species and biodiversity conservation [@williams2002].

@@

There is a natural trade-off between quality and quantity of data that can be collected in any monitoring programme: collecting high quality data in a structured manner is costly and requires trained specialists, thus limiting the amount of data that can be collected, while participatory monitoring, i.e. the collection of ecological data by members of the public [also called citizen or community science, @fraisl2022], allows to greatly reduce costs and extend spatial and taxonomic scales of monitoring at the expense of data quality and risk of bias [@johnston2023]. Consequently, many large-scale monitoring programmes are often limited to presence-absence or very simple count observations, making them suitable for the development of indicators of species distributions and perhaps population trends, but usually not of abundance, population dynamics, and demographic rates (+ example refs). The exception here are monitoring programmes that succeed in making use of a large number of citizen scientists that have been trained to collect data and record metadata in a structured manner and according to a carefully designed protocol. For example, in the United States hunters participate in the collection of bands and wings from harvested American Woodcock (*Scopolax minor)* to estimate survival and age ratios [@zimmerman2010]. <!-- can we think of 1-2 internationally well-known examples? Check out MAPS (Monitoring of Avian Productivity and Survivorship) and CES (Constant Effort Sites) --> In Norway there is a monitoring programme for terrestrial game bird species called "Hønsefuglportalen" (= "game bird portal", <https://honsefugl.nina.no/Innsyn/en>). It is a line transect survey programme carried out annually in \>120 localities across the country (\>2000 transects) by trained volunteers using pointing dogs. The programme has a well developed protocol for recording bird observations, auxiliary data, and relevant metadata (+ref to protocol?) and established routines for quality control and annual releases of publicly available data via GBIF. As such, it is particularly well suited to become part of a workflow for producing and updating abundance and population indicators on an annual basis.

The line transect data from "Hønsefuglportalen" has been used previously for estimating abundance trends of willow ptarmigan (*Lagopus lagopus*) across Norway [e.g. @bowler2020; @nilsen2020] but large-scale estimation of demographic rates underlying these trends has thus far remained as untapped potential of the dataset. @nilsen2023 recently developed a novel integrated distance sampling model (IDSM) which successfully uses the age of idividuals detected along line transects data coupled with radio-telemetry data to estimate not just abundance, but also survival and recruitment across years. In this study, we adapt the model of @nilsen2023 to run not just on a single site but on all areas with publicly available line transect data from "Hønsefuglportalen" simultaneously. Unlike several previous studies applying integrated models for population dynamics to multiple (sub-) populations separately and comparing results [e.g. @robinson2014; @nater2023], we opt for an approach explicitly integrating across space, thus allowing for sharing of information across locations and -- in effect -- space-for-time substitution [e.g. @horswill2019; @morrison2022]. We then apply the resulting multi-area IDSM to "Hønsefuglportalen" data on willow ptarmigan to estimate population size, age-structure, survival, recruitment, and impacts of small rodent abundance across 41 reporting districts and 15 years (2007-2021) for this popular small-game species. We further embed the modelling workflow in a reproducible, semi-automated pipeline that will greatly facilitate the repeated calculation of abundance and population indicators at different spatial scales as new data are added every year.

@@

The willow ptarmigan is a tetraoid bird with a circumpolar distribution, mainly inhabiting sub-alpine and arctic ecosystems [see e.g. @fuglei2020]. While the species is currently listed as Least Concern (LC) both in the global and Norwegian Red List of Species, it has undergone rather dramatic declines in abundance in Norway since the turn of the 20th century [@hjeljord2022]. While the main reason for the long-term decline in abundance remain unresolved, willow ptarmigan are considered sentinel species that are sensitive to both climate change and land use changes [@Henden2017; @Storch2007]. Moreover, being one of only a handful of bird species that spend the winter in mountain ecosystems in Scandinavia, they are important components of the ecosystem as prey species for the gyrfalcon [@Alastair2020] <!-- do we have other "important" species to add here. Arctic fox? Some other endangered beasts perhaps? -->. The willow ptarmigan has a relatively fast pace of life [@sandercock2005; @Steen1996], and can display substantial spatio-temporal variation in demographic rates [@Bowler2020]. Their population dynamics are characterized by large inter-annual fluctuations in abundance [@hjeljord2022], and previous research has tied these fluctuations to rodent cycle through shared predators [@hagen1952; @Bowler2020]. This tight relationship between the breeding success of ground nesting birds and the rodent cycle is known as the Alternative Prey Hypothesis and has been a central part of Fennoscandian research on grouse population dynamics for many decades [@hagen1952; @Linden1988; @Steen1985]. In addition, spring weather conditions and phenology is known to have considerable effects on breeding success and recruitment rates [@Eriksen2023; @Steen1985]. Across their distributional range, they are iconic species with a high cultural value, partly linked to their popularity as game species. The latter means that information about spatio-temporal variation in demographic rates and population dynamics is particularly important in order to design sustainable harvest strategies. In addition, as a sentinel species the willow ptarmigan is well suited as an indicator species for ecosystem status; in Norway it is included in both the main national biodiversity [Nature Index for Norway, @jakobsson2020, <https://www.naturindeks.no/Indicators/lirype>] and ecosystem [Assessment of the Ecological Condition, @framstad2022] assessments.

@@

@@

@@

<!-- move this here but can go either way on location. -->

We used all publicly available data for the period 2007-2021, which included a total of 2225 transects in 41 different reporting districts spanning 9 counties and 50 municipalities. Transects on which no willow ptarmigan were ever observed (i.e. species absence likely due to low habitat suitability) were dropped, resulting in 2077 transects used in the analyses. For the radio-telemetry aspect of the model, we used data from years 2015 - 2020, and the total sample size across these years was 139 birds for the Aug-Jan period and 258 birds for the Feb-Jul period.

@@

The highest population growth rates were estimated for areas with relatively low population densities across latitudes but we did not find evidence for a strong association between population growth rates and population densities overall ([@fig-paramCorrs] A).

@@

Recruitment rates varied between `r R.min$post.sum` (area "`r R.min$area.name`") and `r R.max$post.sum` (area "`r R.max$area.name`") and displayed a spatial pattern opposite to that of annual survival, i.e., lower recruitment rates co-occurring with higher survival rates and vice-versa (@fig-pVRB; @fig-paramCorrs B). Across all areas and years, average recruitment rate was `r R.hyper.mean$post.sum`. Unlike for survival, the model predicted similar magnitudes of spatial and temporal variation (random effect SDs of `r R.sd.spatial$post.sum` and `r R.sd.temporal$post.sum`, respectively), and about twice as much residual variation (`r R.sd.residual$post.sum`).

@@

Detection decay parameters, which determine detection probability in distance sampling surveys, varied across areas from between `r Det.min$post.sum` in "`r Det.min$area.name`" to `r Det.max$post.sum` in "`r Det.max$area.name`, resulting in detection probabilities over the transect sites (truncation distance = 200 m) from `r pDet.min$post.sum` to `r pDet.max$post.sum`, respectively. The global average detection decay was `r Det.hyper.mean$post.sum` (detection probability = `r pDet.hyper.mean$post.sum`), and in general, higher values were more common in the Southern half of the country than the Northern half (SF"Avg_detect_Map.pdf"). Variation in detection over time was modest (SF "TimeSeries_pDetect.pdf"), and among-year variation in detection decay (random effect log SD = `r Det.sd.temporal$post.sum`) was almost identical in magnitude to spatial (`r Det.sd.spatial$post.sum`) and residual (`r Det.sd.residual$post.sum`) variation.

@@

Building on the work of @nilsen2023, we used a structured community science monitoring program and a novel integrated population model to estimate spatial and temporal variation in demography of a culturally important game bird species. Although our study was exploratory in nature, the patterns observed generally conformed to those expected from ecological and life history theory. Apparent trade-offs between survival and recruitment was observed. Support was also found for the benefits of alternative prey to ptarmigan (i.e., alternative prey hypothesis). A muted relationship between population density and growth rates was found, however, this was possibly due to omitted environmental gradients that would have better described the availability of resources across the large spatial extent of the study.

@@

The rich spatial distribution of sampling afforded us the opportunity to explore variation in density and vital rates across a relatively large spatial extent. Density followed the expected pattern of lower density at northern latitudes and western longitudes. Within the spatial extent of our study, these areas experience the most extreme climatic conditions including coldest temperatures and shortest growing season. Ptarmigan are income breeders using food resources acquired from nesting areas to supply energy and nutrients for egg production and incubation [@sandercock2005]. Reduced primary production limits food availability which limits carrying capacity relative to more southern populations.

We did not find a clear relationship between density and growth rates. However, this is unsurprising as it appears to be substantial variation in demography which suggests carrying capacity varies over space. This notion is further supported by the residual variation unaccounted for in the model which is likely environmental conditions that govern resource availability.

[@Wilson_Martin_2011] found greater fecundity at greater latitudes compared to the southern terminus of their distribution in the USA; however, annual survival was less.

@@

The Alternative Prey Hypothesis (APH) has been well-supported throughout the literature across multiple taxa [e.g., @hagen1952; @Kjellander_Nordström_2003; @Reif_Tornberg_Jungell_Korpimäki_2001] and for willow ptarmigan [@bowler2020]. We also found support for APH but the positive effect of rodents on recruitment were more pronounced at northern latitudes while equivocal elsewhere. Predator-prey dynamics are notoriously complex and interact with a multitude of environmental variables. @Nielsen_1999 found that gyrfalcon predation on ptarmigan was greatest during years of low ptarmigan abundance which suggests that the availability of alternative prey is most important at low densities as exhibited at northern latitudes in our study.

@@

Management decisions rarely and shouldn't occur at the resolution of large geo-political boundaries (e.g., Norway) because of the variation in demographic processes that occur. Our study suggests decisions should be spatially refined to capture "local" population processes that will interact with management.

Management also includes optimization of monitoring
